---
title: "EDA"
format: html
editor: visual
---

# EDA

## Our Dataset
*briefly describe the data and the variables you have to work with (just discuss the ones you’ll investigate/use in your analysis).*

## Purpose
*describe the purpose of your EDA and ultimate goal of modeling.*

## Importing the Data + Initial Data Processing

```{r}
#| output: false

# import libraries
library(readr)
library(tidyr)
library(dplyr)
library(tidyverse)
```

How is the data stored?

```{r}
# read in the data
df <- readr::read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
head(df)
```

A lot of the columns could have better names, and many of them contain data that should be stored as factors instead of doubles. Let's update the column names and how the data is stored.

```{r}
# rename columns and update datatypes/data storage
df <- df |>
  rename(Diabetes=Diabetes_binary) |>
  mutate(Diabetes=case_when(
    Diabetes==0 ~ "no diabetes",
    Diabetes==1 ~ "prediabetes or diabetes")) |>
  mutate(Diabetes=as.factor(Diabetes)) |>
  
  rename(High_BP=HighBP) |>
  mutate(High_BP=case_when(
    High_BP==0 ~ "no",
    High_BP==1 ~ "yes")) |>
  mutate(High_BP=as.factor(High_BP)) |>
  
  rename(High_Cholesterol=HighChol) |>
  mutate(High_Cholesterol=case_when(
    High_Cholesterol==0 ~ "no",
    High_Cholesterol==1 ~ "yes")) |>
  mutate(High_Cholesterol=as.factor(High_Cholesterol)) |>
  
  rename(Checked_Cholesterol_5years=CholCheck) |>
  mutate(Checked_Cholesterol_5years=case_when(
    Checked_Cholesterol_5years==0 ~ "no",
    Checked_Cholesterol_5years==1 ~ "yes")) |>
  mutate(Checked_Cholesterol_5years=as.factor(Checked_Cholesterol_5years)) |>
  
  mutate(Smoker=case_when(
    Smoker==0 ~ "no",
    Smoker==1 ~ "yes")) |>
  mutate(Smoker=as.factor(Smoker)) |>
  
  mutate(Stroke=case_when(
    Stroke==0 ~ "no",
    Stroke==1 ~ "yes")) |>
  mutate(Stroke=as.factor(Stroke)) |>
  
  rename(Heart_Disease_or_Attack=HeartDiseaseorAttack) |>
  mutate(Heart_Disease_or_Attack=case_when(
    Heart_Disease_or_Attack==0 ~ "no",
    Heart_Disease_or_Attack==1 ~ "yes")) |>
  mutate(Heart_Disease_or_Attack=as.factor(Heart_Disease_or_Attack)) |>
  
  rename(Physical_Activity_30days=PhysActivity) |>
  mutate(Physical_Activity_30days=case_when(
    Physical_Activity_30days==0 ~ "no",
    Physical_Activity_30days==1 ~ "yes")) |>
  mutate(Physical_Activity_30days=as.factor(Physical_Activity_30days)) |>
  
  rename(Daily_Fruit=Fruits) |>
  mutate(Daily_Fruit=case_when(
    Daily_Fruit==0 ~ "no",
    Daily_Fruit==1 ~ "yes")) |>
  mutate(Daily_Fruit=as.factor(Daily_Fruit)) |>
  
  rename(Daily_Vegetable=Veggies) |>
  mutate(Daily_Vegetable=case_when(
    Daily_Vegetable==0 ~ "no",
    Daily_Vegetable==1 ~ "yes")) |>
  mutate(Daily_Vegetable=as.factor(Daily_Vegetable)) |>
  
  rename(Heavy_Alcohol_Consumption=HvyAlcoholConsump) |>
  mutate(Heavy_Alcohol_Consumption=case_when(
    Heavy_Alcohol_Consumption==0 ~ "no",
    Heavy_Alcohol_Consumption==1 ~ "yes")) |>
  mutate(Heavy_Alcohol_Consumption=as.factor(Heavy_Alcohol_Consumption)) |>
  
  rename(Has_Healthcare=AnyHealthcare) |>
  mutate(Has_Healthcare=case_when(
    Has_Healthcare==0 ~ "no",
    Has_Healthcare==1 ~ "yes")) |>
  mutate(Has_Healthcare=as.factor(Has_Healthcare)) |>
  
  rename(No_Doctor_bc_Cost=NoDocbcCost) |>
  mutate(No_Doctor_bc_Cost=case_when(
    No_Doctor_bc_Cost==0 ~ "no",
    No_Doctor_bc_Cost==1 ~ "yes")) |>
  mutate(No_Doctor_bc_Cost=as.factor(No_Doctor_bc_Cost)) |>
  
  rename(General_Health=GenHlth) |>
  mutate(General_Health=case_when(
    General_Health==1 ~ "excellent",
    General_Health==2 ~ "very good",
    General_Health==3 ~ "good",
    General_Health==4 ~ "fair",
    General_Health==5 ~ "poor")) |>
  mutate(General_Health=as.factor(General_Health)) |>
  
  rename(Bad_Mental_Days_of_Past_30=MentHlth) |>
  rename(Bad_Physical_Days_of_Past_30=PhysHlth) |>
    
  rename(Difficulty_Walking=DiffWalk) |>
  mutate(Difficulty_Walking=case_when(
    Difficulty_Walking==0 ~ "no",
    Difficulty_Walking==1 ~ "yes")) |>
  mutate(Difficulty_Walking=as.factor(Difficulty_Walking)) |>
  
  mutate(Sex=case_when(
    Sex==0 ~ "female",
    Sex==1 ~ "male")) |>
  mutate(Sex=as.factor(Sex)) |>

  mutate(Age=case_when(
    Age==1 ~ "18-24",
    Age==2 ~ "25-29",
    Age==3 ~ "30-34",
    Age==4 ~ "35-39",
    Age==5 ~ "40-44",
    Age==6 ~ "45-49",
    Age==7 ~ "50-54",
    Age==8 ~ "55-59",
    Age==9 ~ "60-64",
    Age==10 ~ "65-69",
    Age==11 ~ "70-74",
    Age==12 ~ "75-79",
    Age==13 ~ "80 or older")) |>
  mutate(Age=as.factor(Age)) |>
  
  rename(Highest_Education_Achieved=Education) |>
  mutate(Highest_Education_Achieved=case_when(
    Highest_Education_Achieved==1 ~ "kindergarten",
    Highest_Education_Achieved==2 ~ "1st-8th grade",
    Highest_Education_Achieved==3 ~ "9th-11th grade",
    Highest_Education_Achieved==4 ~ "12th grade or GED",
    Highest_Education_Achieved==5 ~ "1-3 years college",
    Highest_Education_Achieved==6 ~ "4+ years college")) |>
  mutate(Highest_Education_Achieved=as.factor(Highest_Education_Achieved)) |>

  mutate(Income=case_when(
    Income==1 ~ "less than $10,000",
    Income==2 ~ "less than $15,000",
    Income==3 ~ "less than $20,000",
    Income==4 ~ "less than $25,000",
    Income==5 ~ "less than $35,000",
    Income==6 ~ "less than $50,000",
    Income==7 ~ "less than $75,000",
    Income==8 ~ "more than $75,000")) |>
  mutate(Income=as.factor(Income))

head(df)
```

Now our numeric variables consist of:

- BMI
- Bad_Mental_Days_of_Past_30
- Bad_Physical_Days_of_Past_30

and the rest of the columns contain factor/categorical variables, including our response variable, which is Diabetes.

Now let's explore what values exist in the data and if any cleaning needs to be done.

```{r}
# check for missing values
sum_na <- function(column){
 sum(is.na(column))
}

na_counts <- df |>
 summarize(across(everything(), sum_na))

na_counts
```

There are no missing values! What values do we see in each categorical column, and how does each value relate to the response variable?

## Data Summaries

```{r}
# Contingency tables!
table(df$Diabetes, High_BP=df$High_BP)
table(df$Diabetes, High_Cholesterol=df$High_Cholesterol)
table(df$Diabetes, Checked_Cholesterol_5years=df$Checked_Cholesterol_5years)
table(df$Diabetes, Smoker=df$Smoker)
table(df$Diabetes, Stroke=df$Stroke)
table(df$Diabetes, Heart_Disease_or_Attack=df$Heart_Disease_or_Attack)
table(df$Diabetes, Physical_Activity_30days=df$Physical_Activity_30days)
table(df$Diabetes, Daily_Fruit=df$Daily_Fruit)
table(df$Diabetes, Daily_Vegetable=df$Daily_Vegetable)
table(df$Diabetes, Heavy_Alcohol_Consumption=df$Heavy_Alcohol_Consumption)
table(df$Diabetes, Has_Healthcare=df$Has_Healthcare)
table(df$Diabetes, No_Doctor_bc_Cost=df$No_Doctor_bc_Cost)
table(df$Diabetes, General_Health=df$General_Health)
table(df$Diabetes, Difficulty_Walking=df$Difficulty_Walking)
table(df$Diabetes, Age=df$Age)
table(df$Diabetes, Highest_Education_Achieved=df$Highest_Education_Achieved)
table(df$Diabetes, Income=df$Income)
```

Let's look at each of the categorical variable frequencies again, but this time with plots:

```{r}
# side-by-side bar plots
ggplot(data = df |>
  group_by(Diabetes, High_BP) |>
  summarize(count = n()), aes(x = High_BP, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts with and without High Blood Pressure") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, High_Cholesterol) |>
  summarize(count = n()), aes(x = High_Cholesterol, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts with and without High Cholesterol") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Checked_Cholesterol_5years) |>
  summarize(count = n()), aes(x = Checked_Cholesterol_5years, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Checked_Cholesterol_5years") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Smoker) |>
  summarize(count = n()), aes(x = Smoker, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Smoker") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Stroke) |>
  summarize(count = n()), aes(x = Stroke, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Stroke") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Heart_Disease_or_Attack) |>
  summarize(count = n()), aes(x = Heart_Disease_or_Attack, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Heart_Disease_or_Attack") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Physical_Activity_30days) |>
  summarize(count = n()), aes(x = Physical_Activity_30days, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Physical_Activity_30days") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Daily_Fruit) |>
  summarize(count = n()), aes(x = Daily_Fruit, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Daily_Fruit") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Daily_Vegetable) |>
  summarize(count = n()), aes(x = Daily_Vegetable, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Daily_Vegetable") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Heavy_Alcohol_Consumption) |>
  summarize(count = n()), aes(x = Heavy_Alcohol_Consumption, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Heavy_Alcohol_Consumption") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Has_Healthcare) |>
  summarize(count = n()), aes(x = Has_Healthcare, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Has_Healthcare") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, No_Doctor_bc_Cost) |>
  summarize(count = n()), aes(x = No_Doctor_bc_Cost, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. No_Doctor_bc_Cost") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, General_Health) |>
  summarize(count = n()), aes(x = factor(General_Health, level=c('poor', 'fair', 'good', 'very good', 'excellent')), y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. General_Health") + 
  geom_bar(position = "dodge", stat = "identity") +
  xlab("General_Health") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))

ggplot(data = df |>
  group_by(Diabetes, Difficulty_Walking) |>
  summarize(count = n()), aes(x = Difficulty_Walking, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Difficulty_Walking") + 
  geom_bar(position = "dodge", stat = "identity")

ggplot(data = df |>
  group_by(Diabetes, Age) |>
  summarize(count = n()), aes(x = Age, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Age") + 
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))

ggplot(data = df |>
  group_by(Diabetes, Highest_Education_Achieved) |>
  summarize(count = n()), aes(x = factor(Highest_Education_Achieved, level=c("kindergarten", "1st-8th grade", "9th-11th grade", "12th grade or GED", "1-3 years college", "4+ years college")), y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Highest_Education_Achieved") + 
  geom_bar(position = "dodge", stat = "identity") +
  xlab("Highest_Education_Achieved") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))

ggplot(data = df |>
  group_by(Diabetes, Income) |>
  summarize(count = n()), aes(x = Income, y = count, fill = Diabetes)) +
  ggtitle("Diabetes counts vs. Income") + 
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1))
```

From the proportions of observations in each category with and without diabetes, it looks like the variables that are most interesting to use for prediction of Diabetes are:

- High_BP
- High_Cholesterol
- Smoker
- Stroke
- Heart_Disease_or_Attack
- Physical_Activity_30days
- Daily_Fruit
- Daily_Vegetable
- General_Health
- Difficulty_Walking
- Age (*maybe*)
- Highest_Education_Achieved
- Income

Now let's look at the distribution of numerical values:

```{r}
# density plots
ggplot(df, aes(x = BMI)) + geom_density(aes(fill = Diabetes), alpha=0.5, adjust=2) + ggtitle("BMI by Diabetes")

ggplot(df |> filter(Bad_Mental_Days_of_Past_30>0), aes(x = Bad_Mental_Days_of_Past_30)) + geom_density(aes(fill = Diabetes), alpha=0.5, adjust=3) + ggtitle("Bad_Mental_Days_of_Past_30 by Diabetes")

ggplot(df |> filter(Bad_Physical_Days_of_Past_30>0), aes(x = Bad_Physical_Days_of_Past_30)) + geom_density(aes(fill = Diabetes), alpha=0.5, adjust=3) + ggtitle("Bad_Physical_Days_of_Past_30 by Diabetes")

# box and whisker plots
ggplot(df) + geom_boxplot(aes(x = Diabetes, y = BMI, fill = Diabetes)) + ggtitle("BMI vs. Diabetes")
ggplot(df) + geom_boxplot(aes(x = Diabetes, y = Bad_Mental_Days_of_Past_30, fill = Diabetes)) + ggtitle("Bad_Mental_Days_of_Past_30 vs. Diabetes")
ggplot(df) + geom_boxplot(aes(x = Diabetes, y = Bad_Physical_Days_of_Past_30, fill = Diabetes)) + ggtitle("Bad_Physical_Days_of_Past_30 vs. Diabetes")
```

From these plots, it appears that all 3 numerical variables are correlated with Diabetes and would be useful in prediction.

[Click here for the Modeling Page](Modeling.html)